---
title: "Homework 3"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align="center", message=FALSE, warning=FALSE)
```

**This homework is due on Wednesday, October 22 2025 by 11:59pm. Please show all your work and justify your answers. Use R to compute the probabilities.**

1. Trees are subjected to different levels of carbon dioxide atmosphere with 6\% of them in a minimal growth condition at 350 parts per million (ppm), 10\% at 450 ppm (slow growth), 47\% at 550 ppm (moderate growth), and 37\% at 650 ppm (rapid growth). What are the mean and standard deviation of the carbon dioxide atmosphere (in ppm) for these trees in ppm? 

```{r, echo=TRUE}

```

2. Suppose that in the U.S. in 2011, 70% of new start-ups failed in their first year. If there were 50 start-ups in the DMV area in 2011, what is the probability that 35 or more failed? What assumption did you need to make about the DMV?

```{r, echo=TRUE}

```

3. A research team has developed a face recognition device to match photos in a database. From laboratory tests, the recognition accuracy is 92% and trials are assumed to be independent. 

- If the research team continues to run laboratory tests, what is the mean number of trials until failure? 

- What is the probability that the first failure occurs on the tenth trial? 

- To improve the recognition algorithm, a chief engineer decides to collect 10 failures. How many trials are expected to be needed? 

4. You roll a die until obtaining two consecutive numbers that match. What it is the probability that it takes exactly $x$ rolls to do this, for $x$ an integer greater than 1?

5. The MIT-BIT arrhythmia database contains 48 heart signal recordings with 22 and 26 from female and male patients, respectively. For a classification task, an analyst randomly selects 36 records for predictive model training and keeps the other 12 records for testing the model performance. 

-	What is the expected number of female patient recordings selected for training?

-	What is the probability that at least three male patient recordings are selected for training? 

-	What is the probability that the same number of male and female patients are used for testing the model performance? 

6.	An engineer is participating in a research project on the title patterns of junk emails. The number of junk emails which arrive in an individual’s account every hour follows a Poisson distribution with a mean of 1.2.

-	What is the expected number of junk emails that an individual receives in an 8 hour day? 

-	What is the probability that an individual receives more than two junk emails for the next three hours? 

-	What is the probability that an individual receives no junk email for two hours? 


7. Consider a random variable $X$ with probability density function $f(x) = \theta x^{\theta-1}$ for $0<x<1$, where $\theta>0$ is the parameter that controls the pdf.

- Show that $f(x)$ is a valid pdf

- Plot the pdf for $\theta \in \{1/3,1/2,1,2,3\}$. Describe how the distribution changes for different values of $\theta$.

- Find a general expression for the expected value and variance of $X$.

- Find a general expression for the cumulative distribution function (cdf)

8. Web crawlers need to estimate the frequency of changes to Web sites to maintain a current index for Web searches. Assume that the changes to a Web site follow a Poisson process with a mean of 6 days. Let a random variable $X$ denote the time (in days) until the next change.

- What is the probability that the next change occurs in less than 4.5 days?

- What is the probability that the time until the next change is greater 9.5 days?

- What is the time of the next change that is exceeded with probability 90%?